Transition path sampling based calculations of free energies of enzymatic reactions: the case of human methionine adenosyl transferase and plasmodium vivax adenosine deaminase

### 2022-04-21 tex/manuscript.tex
added some more text

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[]
  {Transition path sampling based calculations of free energies of enzymatic
  reactions: the case of human methionine adenosyl transferase and plasmodium 
  vivax adenosine deaminase}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\abbreviations{TPS,NMR,UV}
\keywords{American Chemical Society, \LaTeX}

@@

We solvated both the systems using TIP3P water molecules 

@@

MD simulations within classical and hybrid quantum mechanics/molecular mechanics 
(QMMM) approximations were carried out using the CHARMM program 
package. \cite{Brooks83JComputChem4p187,Brooks09JComputChem30p1545} 
For the MM calculations the CHARMM36 force field \cite{Brooks09JComputChem30p1545} 
was used and for the QM calculations the approximate PM3 semiempirical 
method \cite{Repasky02JComputChem23p1601} was used. 
For the MAT2A system the partitioned QM region consisted of 
the ATP and MET molecules along with three Mg$^{2+}$ ions 
whereas for the $pv$ADA system it consisted of 
the adenosine molecule, an hydroxide anion, Zn$^{2+}$ ion and 
a portion of the Glu229 residue which was partitioned within the generalized 
hybrid orbital (GHO) scheme. \cite{Gao98JPhysChemA102p4714}

Both the systems were minimized within classical molecular mechanics 
with 50 steps of the
steepest descent method, followed by 2000 steps of the
adopted basis Newton-Raphson method. 
The minimized systems were
heated slowly to 300 K for 35 ps beginning with harmonic
constraints on all atoms except on the H atoms and the TIP3 
waters with gradual reduction of the restraint forces to 1. 
15 ps of equilibration was carried out starting with harmonic 
constraints of force constant 1 followed by 20 ps of contraint free 
equilibration to prepare the systems for TPS simulations. 
During the heating and equilibration steps, bonds that contain 
hydrogen atoms in the MM region were restrained to their equilibrium values 
using the SHAKE procedure. \cite{Ryckaert77JComputPhys23p327}

%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------
TPS simulations require the definition of unique reactant and product states
to collect reactive trajectories based on the Monte Carlo algorithm. Here we 
discuss the details of the reactions involving the MAT2A and the $pv$ADA enzymes 
to identify and define the parameters characterizing the reactant and product basins.

The reaction catalyzed by the human MAT2A enzyme is shown in Fig. \ref{fig:mat2a-reaction}
between adenosyl $5'$-triphosphate (ATP) and L-methionine (MET) molecules resulting in the formation of
S-adenosyl-L-methionine (SAM). SAM is an essential metabolite which is distributed to almost all 
body tissues and fluids, is an universal methyl donor and is of
fundamental importance to metabolism of compounds such as hormones, neureotransmitters, proteins, 
and nucleic acids. \cite{Friedel89Drugs38p389} The reaction was experimentally characterized by 
Firestone et al. \cite{Firestone17JAmChemSoc139p13754} to follow a S$_{\text{N}}2$ transition 
state. The key bond parameters for this reaction consists of the distance between the
nucleophilic sulfur atom of MET and the $5'-C$ atom of the ATP (denoted as $dis(SC)$), and the distance between 
oxygen atom of the phosphate group and the $5'-C$ atom of ATP ($dis(OC)$).
For the TPS calculations we define the reactant state to have $dis(OC)$ to be 
less than 1.7 {\AA} and the product state to have the $dis(SC)$ less than 1.7 {\AA}.

@@

%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
To obtain the free energy profile along the order parameter, we use the method 
discussed in sec. \ref{ssec:algorithm}. The order paratemer for the MAT2A system 
is defined to be $dis(OC)-dis(SC)$ and that for the $pv$ADA is defined as 
$dis(NC)-dis(OC)$. For the MAT2A system, configurations are sampled within 
the range $[-2.0,2.0]$ and this range is divided into 14 overlapping windows 
($0.09$ {\AA} overlap on either side). For the $pv$ADA system configurations are
sampled along the order parameter within the range $[-2.0, 2.0]$ where the range is
divided into 20 overlapping windows ($0.09$ {\AA} overlap on either side of each window). 
A reactive trajectory from the TPS ensemble is chosen to be the guide trajectory 
for sampling within each of the windows. To sample in a particular window $w$, a 
frame from this reactive trajectory is chosen such that the value of the 
order parameter calculated for this frame falls within $w$. Using the shooting algorithm
the momenta are perturbed and dynamics is run for 10 steps forwards and backwards with a 
$1\;fs$ time step to obtain a trajectory with a time period of $20\;fs$.    
Within each window 3000 TPS trajectories are collected using the acceptance criteria 
discussed in sec. \ref{ssec:algorithm}. The equilibrium probability distribution function 
for the order parameter within window $w$ is then calculated as normalized histogram
similar to the WHAM procedure. \cite{Kumar92JComputChem13p1011}
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
%------------------------------------------------------------------------------ 
Human methionine adenosyltransferase (MAT2A) is an enzyme that catalyzes the 
reaction between adenosine triphosphaste and L-methionine which leads to the 
formation of S-Adenosylmethionine. \cite{Firestone17JAmChemSoc139p13754,Niland21Biochem60p791} 
The mechanism of this reaction is experimentally determined to follow a 
SN$_2$ transition state with the sulphur atom of the L-methionine as the 
nucleophile with the triphosphate leaving group.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The document title should be given as usual. Some journals require
%% a running title from the author: this should be supplied as an
%% optional argument to \title.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title[]
  {Transition path sampling based calculations of free energies of enzymatic
  reactions: the case of human methionine adenosyl transferase and plasmodium 
  vivax adenosine deaminase}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Some journals require a list of abbreviations or keywords to be
%% supplied. These should be set up here, and will be printed after
%% the title and author information, if needed.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\abbreviations{TPS,WHAM}
\keywords{American Chemical Society, \LaTeX}

@@

We solvated both the systems using TIP3P water molecules \cite{Jorgensen83JChemPhys79p926}

@@

MD simulations within classical and hybrid quantum mechanics/molecular mechanics 
(QMMM) approximations were carried out using the CHARMM program 
package. \cite{Brooks83JComputChem4p187} 
For the MM calculations the CHARMM36 force field \cite{Brooks09JComputChem30p1545} 
was used and for the QM calculations the approximate PM3 semiempirical 
method \cite{Repasky02JComputChem23p1601} was used. 
For the hybrid QMMM simulations, the MAT2A system was partitioned into a QM
region consisting of the ATP and MET molecules along with three Mg$^{2+}$ ions 
and the MM region consists of the rest of the molecules. 
For the $pv$ADA system the QM region consisted of 
the adenosine molecule, an hydroxide anion, Zn$^{2+}$ ion and 
a portion of the Glu229 residue which was partitioned within the generalized 
hybrid orbital (GHO) scheme. \cite{Gao98JPhysChemA102p4714}

The energies of both the systems were minimized using 50 steps of the 
steepest descent method, followed by 2000 steps of the
adopted basis Newton-Raphson method where only classical molecular mechanics 
was used for the dynamics. 
The minimized systems were
heated slowly to 300 K for 35 ps beginning with harmonic
constraints on all atoms except on the H atoms and the TIP3P 
water molecules with gradual reduction of the restraint forces to 1. 
15 ps of equilibration was carried out starting with harmonic 
constraints of force constant 1 followed by 20 ps of contraint free 
equilibration to prepare the systems for TPS simulations. 
During the heating and equilibration steps, bonds that contain 
hydrogen atoms in the MM region were restrained to their equilibrium values 
using the SHAKE procedure. \cite{Ryckaert77JComputPhys23p327}

%------------------------------------------------------------------------------ 
\subsection{Transition path sampling}
%------------------------------------------------------------------------------
TPS simulations require the definition of unique reactant and product states
to collect reactive trajectories based on the Monte Carlo algorithm. Here we 
discuss the details of the reactions involving the MAT2A and the $pv$ADA enzymes 
to identify and define the order parameters characterizing the reactant and product basins.

The reaction catalyzed by the human MAT2A enzyme is
between adenosyl $5'$-triphosphate (ATP) and L-methionine (MET) molecules resulting in the formation of
S-adenosyl-L-methionine (SAM) which is depicted in Fig. \ref{fig:mat2a-reaction}.
SAM is an essential metabolite which is distributed to almost all 
body tissues and fluids, is an universal methyl donor and is of
fundamental importance to the metabolism of compounds such as hormones, neureotransmitters, proteins, 
and nucleic acids. \cite{Friedel89Drugs38p389} This reaction was experimentally characterized by 
Firestone et al. \cite{Firestone17JAmChemSoc139p13754} to follow a S$_{\text{N}}2$ transition 
state. The key bond parameters for this reaction consists of the distance between the
nucleophilic sulfur atom of MET and the $5'-C$ atom of the ATP (denoted as $dis(SC)$), and the distance between 
oxygen atom of the phosphate group and the $5'-C$ atom of ATP ($dis(OC)$).
For the TPS calculations we define the reactant state to have $dis(OC)$ to be 
less than 1.7 {\AA} and the product state to have the $dis(SC)$ less than 1.7 {\AA}.

@@

%------------------------------------------------------------------------------ 
\subsection{Free energy calculations}
%------------------------------------------------------------------------------ 
To obtain the free energy profile along the order parameter, we use the method 
discussed in sec. \ref{ssec:algorithm}. The order paratemer for the MAT2A system 
is defined to be $dis(OC)-dis(SC)$ and that for the $pv$ADA is defined as 
$dis(NC)-dis(OC)$. For the MAT2A system, configurations are sampled within 
the range $[-2.0,2.0]$ and this range is divided into 14 overlapping windows 
($0.09$ {\AA} overlap on either side). 
Configurations are sampled for the $pv$ADA system within the interval $[-3.0, 3.0]$ {\AA}
of the order parameter value. 
This interval is divided into 20 overlapping windows ($0.045$ {\AA} between 
neighboring windows). A reactive trajectory from the TPS ensemble is chosen to be the guide trajectory 
for sampling within each of the windows. To sample in a particular window $w$, a 
frame from this reactive trajectory is chosen such that the value of its 
order parameter falls within $w$. Using the shooting algorithm
the momenta are perturbed and dynamics is run for 10 steps forwards and backwards with a 
$1\;fs$ time step to obtain a trajectory with a time period of $20\;fs$.    
Within each window 3000 TPS trajectories are collected. The 
probability distribution function for the order parameter within window $w$ 
is then calculated as normalized histograms which represent the populations 
of the configurations within the window at equilibrium. 
Then we calculate the free energy profile in a similar fashion to the WHAM 
procedure \cite{Kumar92JComputChem13p1011} which is generally used to obtain the 
potential of mean force from umbrella sampling calculations. 
%------------------------------------------------------------------------------ 
\section{Results and discussion}
%------------------------------------------------------------------------------ 
\subsection{Human Methionine Adenosyltransferase 2A}
%------------------------------------------------------------------------------ 
Human methionine adenosyltransferase (MAT2A) is an enzyme that catalyzes the 
reaction between adenosine triphosphaste and L-methionine which leads to the 
formation of S-Adenosylmethionine. \cite{Firestone17JAmChemSoc139p13754,Niland21Biochem60p791} 
The mechanism of this reaction is experimentally determined to follow a 
SN$_2$ transition state with the sulphur atom of the L-methionine as the 
nucleophile with the triphosphate leaving group.
